---
title: "Analysis of miles per gallon by vehicle transmission type"
output:
  html_document: default
  pdf_document:
    keep_tex: yes
---
```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=8, fig.height=6, fig.path='Figs/',
                      echo=TRUE, warning=FALSE, message=FALSE)
```
```{r echo=FALSE}
suppressPackageStartupMessages({
    library(lattice) 
    library(stargazer)
    library(broom)
    library(knitr)
})

data(mtcars)
```
### Summary

You work for Motor Trend, a magazine about the automobile industry. Looking at a 
data set of a collection of cars, they are interested in exploring the relationship 
between a set of variables and miles per gallon (MPG) (outcome). They are particularly 
interested in the following two questions:

- "Is an automatic or manual transmission better for MPG"
- "Quantify the MPG difference between automatic and manual transmissions"

### Analysis

The data used for the analysis comes from the mtcars data set provided from the 
1974 Motor Trend US magazine, and consists of 32 observations using 11 variables,
see Appendix 1 for data structure.  

We conduct a preliminary exploration of the relationship of MPG to transmission, 
taking into account weight and number of cylinders, which are also generally
related to displacement. A plot (see Figure 1 in the first Appendix) reveals clear 
distinctions in the 4 and 6 cylinder cars, but that it is only 4 cylinder cars 
that markedly show a preference for manual transmission to deliver a better MPG 
performance. The 8 cylinder cars show a preference for automatic transmission for 
MPG performance, but an investigation of the two manual cars in the group reveal 
that they are high performance sports cars, whereas most of the 8 cylinder 
automatics are large family cars or mid-range performance sports cars. 

```{r MTCARSPLOT, eval=FALSE, echo=FALSE}
xyplot(mpg ~ wt | factor(cyl), data=mtcars, pch=19,  
       main="", xlab="Weight / 1,000lb",  ylab="MPG", layout=c(3,1),
       groups = factor(am,labels=c("Auto","Manual")),
       type=c("p","g","r"),auto.key = TRUE)
```
In order to fairly evaluate the efficiency of manual vs. automatic transmission 
we would need to have identical models of cars with the alternative transmissions
driven under controlled conditions to get the best mileage per gallon. We do not 
have such a set of data, so the best we can do is to identify what other factors 
might contribute to a better MPG performance, and try to account for them when 
modelling the effect of transmission on MPG. 

We have already identified number of cylinders and weight as determining factors. 
Therefore rather than using a simple MPG to transmission relationship we should 
try to find a more complete set of variables in order to establish the most 
accurate and parsimonious model to quantify the relationship between MPG and 
transmission.

For this we can conduct a multivariate regression analysis. 

To assist with visualizing the effects of all of the factors we add a column 
hp.wt to the data to show power to weight ratio (hp/wt), which together with 
quarter mile performance (qsec) can assist in understanding performance. In 
Appendix 2 we have printed lists of cars based on transmission type and sorted 
by MPG and performance.

```{r TESTCARS, echo=FALSE}
testCars <- mtcars
testCars <- testCars[order(-testCars$am, -testCars$mpg),]
testCars$am <- factor(testCars$am, labels=c('Automatic','Manual'))
testCars$cyl <- factor(testCars$cyl)
testCars$vs <- factor(testCars$vs, labels=c('V-engine','Straight-engine'))
testCars$gear <- factor(testCars$gear)
testCars$carb <- factor(testCars$carb)
testCars$hp.wt <- testCars$hp / testCars$wt

testCars$vs <- relevel(testCars$vs, "Straight-engine")
testCars$am <- relevel(testCars$am, "Manual")

carMed <- subset(testCars, ! rownames(testCars) == "Maserati Bora")
```
As a comparative starting point we perform a simple evaluation by testing a two sided 
hypothesis that MPG from manual transmissions is better than from automatics, by 
supplying manual and auto transmission data as two samples to the R function 
**t.test** (default 95% confidence level):

```{r echo=FALSE}
autoCars <- mtcars[mtcars$am == 0,]; 
manualCars <- mtcars[mtcars$am == 1,]
tTest <- t.test(manualCars$mpg, autoCars$mpg)
tidyTest <- tidy(tTest)
newNames <- c("diff in means", "mean Manual", "mean Auto", "Test Statistic", 
              "P value", "DF", "low conf", "high conf")
names(tidyTest) <- newNames
tidyTest$`mean Manual` <- round(tidyTest$`mean Manual`, 3)
tidyTest$`mean Auto` <- round(tidyTest$`mean Auto`, 3)
tidyTest$`diff in means` <- tidyTest$`mean Manual` - tidyTest$`mean Auto`
tidyTest$`Test Statistic` <- round(tidyTest$`Test Statistic`, 3)
tidyTest$`P value` <- round(tidyTest$`P value`, 5)
tidyTest$`DF` <- round(tidyTest$`DF`, 2)
tidyTest$`low conf` <- round(tidyTest$`low conf`, 3)
tidyTest$`high conf` <- round(tidyTest$`high conf`, 3)

kable(tidyTest)
```
  
Our simple model's prediction for MPG improvement between manual vs automatic 
transmission is the difference in the means, so we can predict an average 
improvement of **`r tidyTest$"diff in means"` MPG** for manual transmission cars.

However this test ignores various other variables, so to account for sufficient 
of the variables involved in the MPG performance we need to perform some multivariate
tests. We can use the step function of R to test all of the possible useful variations,
and after evaluating the results suggest the best model. See Appendix 3 for results:

```{r MODEL, eval=FALSE, echo=FALSE}
model <- lm(mpg ~ ., data=testCars)
stepmodel <- step(lm(mpg ~ ., data=testCars), trace=0, steps=10000, direction="both")
altmodel <- lm(mpg ~ cyl + hp + wt + qsec + am, data=testCars)
medmodel <- step(lm(data=carMed, mpg ~ .), trace=0, steps=1000, direction="both")

stargazer(model, stepmodel, altmodel, medmodel, type = "html", single.row=TRUE, header=FALSE, 
          object.names = TRUE, report = "v*cs*", intercept.bottom = FALSE, align=TRUE
          ,covariate.labels=c("(Intercept)", "6 cylinders", "8 cylinders", 
                              "Displacement", "Gross horsepower", "Rear axle ratio", 
                              "Weight", "1/4 mile time", "V engine configuration",
                              "Automatic transmission", "4 forward gears",
                              "5 forward gears", "2 carburetors", "3 carburetors", 
                              "4 carburetors", "6 carburetors", "8 carburetors", 
                              "Power to weight ratio")
          )
```

```{r PLOTSMODEL, eval=FALSE, echo=FALSE}
par(mfrow = c(2,2))
plot(stepmodel, id.n = 8)
```
```{r PLOTAMODEL, eval=FALSE, echo=FALSE}
par(mfrow = c(2,2))
plot(altmodel, id.n = 8)
```
```{r PLOTMMODEL, eval=FALSE, echo=FALSE}
par(mfrow = c(2,2))
plot(medmodel, id.n = 8)
```
   
#### Step model  
The suggested variables returned by step were cylinders, horsepower, weight, and 
transmission. 

#### Alternative Model
After analysis of the results we re-introduced i/4 mile time (qsec)
to test as an alternative model, as it seems help account for effects of high 
performance vehicles. 

The adjusted R squared score of the alternative model was only 0.01 better than 
the suggested step model, but as the estimate of transmission effect was **55% higher** 
and seems likely more accurate, it seems a better model. Comparisons of the 
residuals plots in Appendix 3 seem to indicate a better fit for our alternative 
model, the QQ plot is aligned along more points compared to the suggested model, 
the scale location regression line is flatter, and the residuals vs leverage 
regression line is  flatter and centered closer to the zero intercept.

#### Discarded variables
Many of the variables discarded by testing through step seemed to have significant
effects, but were also variable in tandem with the key values of horsepower,
weight and 1/4 mile time. For instance rear axle ratio (drat) was given a high
score, but it seems to be confounded with the simultaneous low horsepower and weight 
of the Honda Civic, which also has the highest rear axle ratio of all of the cars, 
and achieves very good MPG. Therefore drat is noise, and is removed from from the 
model. Likewise for numbers of carburetors, they achieve a number of significant
looking scores but are really linked to the performance of a vehicle better 
measured via horsepower.

#### Medium Model
Some of the residuals exert significant leverage, the Maserati Bora has the highest
performance specification of the manual transmission cars, while the Toyota Corona 
has the worst MPG of the 4 cylinder cars (though still better than any 6 or 8 cylinder
cars). 

If we take out just the Maserati Bora from the data set then step produces what 
seems a more balanced model based just on weight, 1/4 mile time, and transmission
and shows a **further trend** towards manual transmission being better than automatic
transmission - column 4 of the comparative summary in Appendix 3.
</br>
  
### Conclusion and MPG prediction
We have seen that a simple comparison of mean MPG between manual and automatic
transmissions produces a rather high value for the predicted MPG effect, and 
have concluded that a multivariate regression analysis is required to take into
account other variables beside transmission. We have taken the suggested variables
which the step function determined as our first model, as an alternative we have 
added 1/4 mile time to try to better account for high performance vehicles, and 
finally we created a model that eliminates the extremely powerful Maserati Bora
from the data. These 3 models suggest different values for MPG improvement from
manual over automatic transmission, and its possible that the last model is most 
representative of the true difference between manual and automatic transmissions.

The range of possibilities available in the multivariate comparisons suggest an 
improvement in MPG performance for manual transmissions from between 1.8 to 3.2
MPG, with **2.8** as the most likely choice considering all of the vehicles, but
**3.2** as most representative if the high performance Maserati Bora is excluded.

### Appendix 1 - Structure of mtcars table and MPG vs weight analysis

Field     | Description             | Field     | Description      
--------- | ------------            | --------- | ------------
1.  mpg   | Miles/(US) gallon       | 7.  qsec  | 1/4 mile time
2.  cyl   | Number of cylinders     | 8.  vs    | V/S (0 = V, 1 = Straight configuration)
3.  disp  | Displacement (cu.in.)   | 9.  am    | Transmission (0 = automatic, 1 = manual)
4.  hp    | Gross horsepower        | 10. gear  | Number of forward gears
5.  drat  | Rear axle ratio         | 11. carb  | Number of carburetors
6.  wt    | Weight (lb/1000)        |           |                          .

#### Figure 1 - MPG vs Weight

```{r MPGVWT, ref.label="MTCARSPLOT", echo=FALSE, fig.height = 5, fig.width = 9, fig.align = 'center'}
```

### Appendix 2 - data to assist with comparitive analysis
   
#### Figure 2 - Manual and Automatic cars

```{r VTABLE, echo=FALSE, cache=TRUE}
manual=subset(mtcars, am==1)
manual$hp.wt <- manual$hp / manual$wt
manual <- manual[order(-manual$mpg, manual$cyl, manual$hp.wt, manual$qsec),]
kable(manual, caption = "Manual Transmission")

auto=subset(mtcars, am==0)
auto$hp.wt <- auto$hp / auto$wt
auto <- auto[order(-auto$mpg, auto$cyl, auto$hp.wt, auto$qsec),]
kable(auto, caption = "Automatic Transmission")
```
  
</br> 
  
### Appendix 3 - modelling the effects of variables on MPG
</br> 

#### Figure 3 - comparisons of model variables

Model Name  | Variables               | Description
----------- | ------------            | --------- 
model       | lm(mpg ~ ., data=testCars) | Test all variables
stepmodel   | cyl + hp + wt + qsec + am  | Variables of the step model
altmodel    | cyl + hp + wt + qsec + am  | Variables of the alternative model 
medmodel    | wt + qsec + am             | Variables of the medium model 

```{r ref.label="MODEL", echo=FALSE, results="asis"}

```
</br>  
  
#### Figure 4 - Residuals plots from the step model
```{r ref.label="PLOTSMODEL", echo=FALSE, fig.height = 7, fig.width = 9}

```
  
#### Figure 5 - Residuals plots from the alternative model, which includes qsec:
```{r ref.label="PLOTAMODEL", echo=FALSE, fig.height = 7, fig.width = 9}

```
  
The regression table in appendix 3 was created using the stargazer library: http://CRAN.R-project.org/package=stargazer

This document was created using R markdown and knitr to create an HTML file, which 
was then converted to PDF. For reasons of space most of the R code is not reproduced 
here, but the original R markdown file with the embedded R code is available - 
accompanying documentation will provide details of where to obtain it.
